---
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve"]
---

<!-- README.md is generated from README.Rmd. Please edit that file, then
     run devtools::build_readme(). CI (readme.yaml) re-renders it and fails
     on any difference, so the output below cannot drift from the code:
     it once printed a digest that was not SHA-256 of its input (#36). -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# zucrypt

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/pedrobtz/zucrypt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pedrobtz/zucrypt/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/pedrobtz/zucrypt/main/.github/badges/coverage.svg)](https://github.com/pedrobtz/zucrypt/actions/workflows/coverage.yaml)
<!-- badges: end -->

zucrypt is a narrow native cryptographic foundation for the `zu*` package
family: message digests, HMAC, unauthenticated AES-CBC, constant-time
comparison and secure erasure, over raw vectors. The backend is a vendored,
pinned subset of [TF-PSA-Crypto](https://github.com/Mbed-TLS/TF-PSA-Crypto),
the Mbed TLS project's cryptography library — 109 files of it, configured down
to the algorithms above and nothing else. Installing the package needs a C99
compiler and nothing else: no system cryptographic library, no Java, no
Python, and no network access during installation.

The same primitives are published to other packages twice, because the family
consumes siblings in two different ways: as a registered C function table for
consumers that can carry an `Imports:`, and as a static archive
(`libzucrypt.a`) for consumers that cannot.

It is **not** a replacement for
[openssl](https://cran.r-project.org/package=openssl) or
[sodium](https://cran.r-project.org/package=sodium). Those cover far more
ground, and a package that wants broad cryptographic functionality should use
them. zucrypt exists because the `zu*` packages need a small, fixed set of
primitives with a stable C ABI and an installation that depends on nothing.

Deliberately out of scope for now: authenticated encryption, password-based key
derivation, random number generation, signatures and key serialization. Each
needs a concrete consumer before it is added. Office/Excel decryption lives in
[zuxlsx](https://github.com/pedrobtz/zuxlsx), which is the first consumer of
this package, not here.

## Usage

Bytes in, bytes out. Raw vectors only — a character value is never guessed at,
never treated as a file name, and never given an encoding on your behalf.

```{r example}
library(zucrypt)

digest <- crypt_hash(charToRaw("the quick brown fox"))
digest

# Hex is an explicit conversion at the call site, not a default.
paste(format(digest), collapse = "")

# A keyed digest, and the right way to check one.
key <- as.raw(rep(0x0b, 32))
tag <- crypt_hmac(charToRaw("Hi There"), key)
crypt_equal(tag, crypt_hmac(charToRaw("Hi There"), key))
```

Use `crypt_equal()` rather than `identical()` whenever one side is a secret:
an ordinary comparison stops at the first differing byte, so how long it takes
measures how much of the expected value an attacker has guessed.

`crypt_info()` reports what the installed build actually contains, read from
the compiled library rather than from anything written down in R:

```{r info}
info <- crypt_info()
info$algorithms
info$vendored
```

The AES-CBC functions exist too, and are deliberately not shown here: they add
no padding and no authentication, and presenting them beside a hash would
suggest they are a general-purpose way to encrypt something. See
`?crypt_aes_cbc` before using them.

[Getting started](https://pedrobtz.github.io/zucrypt/articles/zucrypt.html)
walks through every function, including what using AES-CBC correctly takes
and what each error means.

## Consuming it from C

Other packages can use these primitives without going through R, in either of
the two ways the `zu*` family links siblings:

* **The static archive** `libzucrypt.a`, for a package that cannot carry an
  `Imports:`. It is installed to `lib/` plus the R sub-architecture
  (`lib/x64/` on Windows), beside the licence of the backend compiled into it.
  This is the primary shape. It is *provisional* until its first consumer —
  `zuxlsx`'s decryption of password-protected workbooks — has linked it, and
  is frozen as ABI 1 in 0.2.0.
* **A registered function table**, for a package that can carry an
  `Imports:`. *Experimental* until a package other than a test fixture uses
  it.

See `?zucrypt_c_api` for both, and for what is and is not promised.

## Status

Version 0.1.0. The six R functions are stable; the C interface is
provisional (the archive) or experimental (the table), as above. Every change
to either is recorded in `NEWS.md`. The design is in `.agents/design.md`, and
`.agents/roadmap.md` records how it was built, what was deliberately left out,
and what comes next.

## Installation

``` r
# install.packages("pak")
pak::pak("pedrobtz/zucrypt")
```
