---
title: "Getting started with zucrypt"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zucrypt)
```

zucrypt has six functions:

| Function | What it is for |
|---|---|
| `crypt_hash()` | A digest of some bytes: SHA-1, SHA-256, SHA-384 or SHA-512 |
| `crypt_hmac()` | A keyed digest (HMAC) of some bytes |
| `crypt_equal()` | Comparing a secret value in constant time |
| `crypt_info()` | What the installed build contains |
| `crypt_aes_cbc_encrypt()`, `crypt_aes_cbc_decrypt()` | Unauthenticated AES-CBC, for formats that specify it |

This article walks through each of them. The short version is that there is
not much to learn, on purpose: every function takes raw vectors and returns
raw vectors, algorithm names are matched exactly, and anything wrong is an
error with a class you can catch.

## Bytes in, bytes out

Every argument that holds data or a key is a **raw vector**. zucrypt never
guesses how to turn something else into bytes, so a character string is an
error rather than a silent conversion:

```{r}
e <- tryCatch(crypt_hash("the quick brown fox"), error = function(e) e)
class(e)
conditionMessage(e)
```

The conversion is always yours to make, and it is always one line:

```{r}
crypt_hash(charToRaw("the quick brown fox"))
```

The same goes for anything else you want to hash. Each of these is a
deliberate choice about *which* bytes, which is exactly the choice a package
should not make for you:

```{r}
# An R object: serialize() decides the bytes, and its `version` and `ascii`
# arguments change them, so pin what you rely on.
digest_of_object <- crypt_hash(serialize(mtcars, NULL, version = 3))

# A file's contents, read as bytes, never as text.
path <- tempfile()
writeLines("some file contents", path)
digest_of_file <- crypt_hash(readBin(path, "raw", n = file.size(path)))
```

A string is text in some encoding. `charToRaw()` gives you the bytes R holds
for it; if the string might not be UTF-8, convert it first with
`enc2utf8()` so that the same text always hashes the same way.

## `crypt_hash()`: digests

The default is SHA-256, and unless an external format requires something
else, that is the right choice:

```{r}
msg <- charToRaw("abc")
crypt_hash(msg)
```

Hex is an explicit step at the call site, not a default:

```{r}
hex <- function(x) paste(format(x), collapse = "")
hex(crypt_hash(msg))
```

That is the FIPS 180-2 test vector for SHA-256("abc"). The other algorithms
are named exactly as `crypt_info()$algorithms` lists them:

```{r}
crypt_info()$algorithms

for (a in crypt_info()$algorithms) {
  cat(sprintf("%-6s %s\n", a, hex(crypt_hash(msg, a))))
}
```

Exact means exact: there is no partial matching and no fallback, so a typo
cannot quietly select a different digest.

```{r}
e <- tryCatch(crypt_hash(msg, "sha"), error = function(e) e)
class(e)
```

SHA-1 is there because some existing formats specify it. Do not choose it for
anything new.

The digest of empty input is well defined, and you get it:

```{r}
hex(crypt_hash(raw(0)))
```

## `crypt_hmac()`: keyed digests

An HMAC is a digest that only someone holding the key can compute, which makes
it the standard way to check that data has not been changed by someone who
does not hold the key. The key is raw too, of any length:

```{r}
key <- charToRaw("Jefe")
tag <- crypt_hmac(charToRaw("what do ya want for nothing?"), key)
hex(tag)
```

That is test case 2 of RFC 4231. The algorithm argument works as it does for
`crypt_hash()`:

```{r}
hex(crypt_hmac(charToRaw("what do ya want for nothing?"), key, "sha512"))
```

In real use the key should be random bytes from a cryptographic source, at
least as long as the digest, and kept secret. zucrypt does not generate
randomness; `openssl::rand_bytes()` or `sodium::random()` do. R's own
`sample()` and `runif()` do not, whatever the seed.

## `crypt_equal()`: checking a secret

When you check a tag you received against the one you expect, compare with
`crypt_equal()`, never `identical()` or `==`:

```{r}
received <- tag
crypt_equal(received, crypt_hmac(charToRaw("what do ya want for nothing?"), key))

tampered <- crypt_hmac(charToRaw("what do ya want for nothing!"), key)
crypt_equal(tampered, tag)
```

An ordinary comparison stops at the first byte that differs, so how long it
takes tells an attacker how many leading bytes of their guess were right —
and a tag can then be guessed one byte at a time. `crypt_equal()` takes the
same time whatever the contents.

The length is not treated as secret: inputs of different lengths are simply
not equal.

```{r}
crypt_equal(tag, tag[1:16])
```

## `crypt_info()`: what is installed

`crypt_info()` reports what this build actually contains. Everything in it is
read from the compiled library, so it describes the code you are running, not
a file that says what was meant to be compiled:

```{r}
str(crypt_info())
```

`vendored` names the bundled cryptography library and its version. That
matters because a vendored library is updated by updating the package, not by
updating your operating system. `abi_version` is the version of the C
interface that other packages link against (see `?zucrypt_c_api`).

## AES-CBC: for formats that specify it

`crypt_aes_cbc_encrypt()` and `crypt_aes_cbc_decrypt()` are **interoperability
tools**, for reading and writing file formats that are defined in terms of
AES-CBC. They are deliberately low-level:

- **No padding is added or removed.** The data must already be a multiple of
  16 bytes.
- **Nothing is authenticated.** Anyone who can reach a ciphertext can change
  it, and decryption will return changed plaintext without complaint.
- The key must be 16, 24 or 32 bytes (AES-128, -192 or -256), and the IV
  exactly 16.

If you want to encrypt something of your own, use a construction that
authenticates — `sodium::data_encrypt()`, for example — rather than
assembling one from these. The rest of this section shows what these
functions do and what "doing it correctly" requires, which is also the
argument for not doing it yourself.

### Reproducing a published vector

This is the first two blocks of the NIST SP 800-38A F.2.1 example, which is
the kind of thing these functions exist for: matching bytes that another
implementation specifies.

```{r}
unhex <- function(x) {
  as.raw(strtoi(substring(x, seq(1, nchar(x), 2), seq(2, nchar(x), 2)), 16L))
}
key <- unhex("2b7e151628aed2a6abf7158809cf4f3c")
iv  <- unhex("000102030405060708090a0b0c0d0e0f")
pt  <- unhex(paste0("6bc1bee22e409f96e93d7e117393172a",
                    "ae2d8a571e03ac9c9eb76fac45af8e51"))

ct <- crypt_aes_cbc_encrypt(pt, key, iv)
hex(ct)
identical(crypt_aes_cbc_decrypt(ct, key, iv), pt)
```

The IV is fixed here only so that the output can be compared with the
standard. An IV used to encrypt real data must be unpredictable and never
reused with the same key.

### Padding is yours

Data that is not a multiple of 16 bytes is refused, not padded:

```{r}
e <- tryCatch(crypt_aes_cbc_encrypt(charToRaw("hello"), key, iv),
              error = function(e) e)
class(e)
```

If your format uses PKCS#7 padding, apply and check it yourself:

```{r}
pad_pkcs7 <- function(x) {
  n <- 16L - length(x) %% 16L
  c(x, as.raw(rep(n, n)))
}
unpad_pkcs7 <- function(x) {
  n <- as.integer(x[length(x)])
  stopifnot(n >= 1L, n <= 16L, all(x[length(x) - seq_len(n) + 1L] == as.raw(n)))
  x[seq_len(length(x) - n)]
}

padded <- pad_pkcs7(charToRaw("hello"))
length(padded)
rawToChar(unpad_pkcs7(crypt_aes_cbc_decrypt(
  crypt_aes_cbc_encrypt(padded, key, iv), key, iv)))
```

### Formats that restart the chain

Some formats split data into segments and restart CBC at each boundary with
its own IV — Office's agile encryption does this in 4096-byte segments. Each
call starts from the IV it is given, so a segmented format is one call per
segment:

```{r}
# The NIST plaintext above as two 16-byte "segments", each from the same IV.
segments <- split(pt, rep(1:2, each = 16))
segmented <- unlist(lapply(segments, crypt_aes_cbc_encrypt, key = key, iv = iv),
                    use.names = FALSE)

# The first segment is the first block of the stream, so it matches:
identical(segmented[1:16], ct[1:16])
# The second does not continue the chain; it restarted from the IV:
identical(segmented[17:32], ct[17:32])
identical(segmented[17:32], crypt_aes_cbc_encrypt(pt[17:32], key, iv))
```

In a real format each segment gets its own IV, derived as the format
specifies; here both use one IV only to keep the example short.

### Authenticating what you encrypt

If you must build on CBC, the minimum is **encrypt-then-MAC**: compute an HMAC
over the IV and the ciphertext with a *second, independent* key, send the tag
alongside, and check it with `crypt_equal()` *before* decrypting anything.

```{r}
enc_key <- unhex("603deb1015ca71be2b73aef0857d77811f352c073b6108d72d9810a30914dff4")
mac_key <- as.raw(1:32)   # in real use: 32 random bytes, kept secret

seal <- function(plaintext, iv) {
  ct <- crypt_aes_cbc_encrypt(pad_pkcs7(plaintext), enc_key, iv)
  list(iv = iv, ct = ct, tag = crypt_hmac(c(iv, ct), mac_key))
}
open_sealed <- function(box) {
  expected <- crypt_hmac(c(box$iv, box$ct), mac_key)
  if (!crypt_equal(box$tag, expected)) stop("authentication failed")
  unpad_pkcs7(crypt_aes_cbc_decrypt(box$ct, enc_key, box$iv))
}

box <- seal(charToRaw("meet at noon"), iv)
rawToChar(open_sealed(box))

box$ct[1] <- xor(box$ct[1], as.raw(1))   # one bit flipped in transit
tryCatch(open_sealed(box), error = function(e) conditionMessage(e))
```

Without the tag, that flipped bit would have decrypted into altered plaintext
and nothing would have noticed. Getting every detail of this right — separate
keys, the IV inside the MAC, checking before decrypting, a fresh random IV each
time — is exactly why a ready-made authenticated construction is the better
tool for your own data.

## When something goes wrong

Every error is a condition with a specific class, then `zucrypt_error`, then
`error`. Branch on the class, never on the message, which may be reworded:

```{r}
e <- tryCatch(crypt_aes_cbc_encrypt(raw(32), raw(15), raw(16)),
              error = function(e) e)
class(e)
e$algorithm
```

| Class | Raised when |
|---|---|
| `zucrypt_invalid_argument` | An argument has the wrong type, for example a string where raw bytes are needed |
| `zucrypt_unsupported_algorithm` | An algorithm name is not one of `crypt_info()$algorithms` |
| `zucrypt_bad_length` | A key, IV or data length is not allowed |
| `zucrypt_memory_error` | An allocation failed |
| `zucrypt_backend_error`, `zucrypt_internal_error` | The cryptographic library refused, or an invariant broke; please report these |

Every condition also carries `algorithm` and `native_status` fields. None of
them ever contains a key, an IV or any of your data, so an error message is
safe to paste into a bug report.

```{r}
handled <- tryCatch(
  crypt_hash(msg, "md5"),
  zucrypt_unsupported_algorithm = function(e) "md5 is not provided; use sha256"
)
handled
```

## From C

The same primitives are available to other packages' compiled code, either
through a registered function table or by linking a static library, with no
R involved at run time. See `?zucrypt_c_api`.

## Is this the right package?

zucrypt is deliberately narrow. For most R work another package is the better
choice:

| You want | Reach for |
|---|---|
| Keys, certificates, signatures, broad algorithm coverage | [openssl](https://cran.r-project.org/package=openssl) |
| Modern authenticated encryption that is hard to misuse | [sodium](https://cran.r-project.org/package=sodium) |
| A cache key for an R object, or a fast non-cryptographic hash | [digest](https://cran.r-project.org/package=digest) |
| SHA-2 and HMAC with no system library, a constant-time compare, or cryptography callable from your package's C code | zucrypt |
